### 2017-02-24 2017mergingmerges/section/01-Introduction.tex
Fixing start of S2, also pagination

We might instead define some sort of uniform interface for data sources, with a single `pull' function that provides the next value in each stream. Each operator could be given this interface, so that the next value from each result stream is computed on demand. This is approach is commonly taken with implementations of physical operators in data base systems. However, this `pull only' model does not support operators with multiple outputs, such as our derived @uniquesUnion@ operator, at least not without unbounded buffering. If a consumer pulls many elements from the result @sUnique@ stream, then the implementation needs to pull the corresponding source elements from @sIn1@ \emph{as well} as buffering an arbitrary number of matching elements from @sIn2@. It needs to buffer an aribrary number of elements from @sIn2@ because there is no guarantee of when a consumer will also pull from the @sUnion@ result stream. Once that happens the elements from @sIn2@ no longer need to be retained, but not before.

@@

Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse our example because they do not support operators with multiple inputs. We refer to such a situation as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow graph. Some systems support both pull and push; data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators. More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.

@@

Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent operators such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}. The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level and, other than conservative extensions, only supports affine loops, which rules out operators such as @group@ and @merge@.

In our new system we still view the program as a concurrent process network, where each operator is a separate process, and the stream data flows through communication channels between the processes. Each operator is expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream. The fusion transform then takes the concurrent process network and \emph{sequentializes} it into a single process by choosing a particular evaluation order that requires no unbounded intermediate buffers. When the fusion transformation succeeds we know it has worked. There is no need to inspect intermediate representations of the compiler to debug poor performance, which is a common problem in systems based on general purpose program transformations \cite{lippmeier2012:guiding}.

>>>>

We might instead define some sort of uniform interface for data sources, with a single `pull' function that provides the next value in each stream. Each operator could be given this interface, so that the next value from each result stream is computed on demand. This is approach is commonly taken with implementations of physical operators in data base systems. However, this `pull only' model does not support operators with multiple outputs, such as our derived @uniquesUnion@ operator, at least not without unbounded buffering. Suppose a consumer pulls many elements from the result @sUnique@ stream. The implementation needs to pull the corresponding source elements from @sIn1@ \emph{as well} as buffering an arbitrary number of matching elements from @sIn2@. It needs to buffer an aribrary number of elements from @sIn2@ because there is no guarantee of when a consumer will also pull from the @sUnion@ result stream. Once that happens the elements from @sIn2@ no longer need to be retained, but not before.

@@

Push-based systems such as foldr/build fusion~\cite{gill1993short} also cannot fuse our example because they do not support operators with multiple inputs. We refer to such a situation as a \emph{join} --- in our example the @merge@ operator expresses a join in the data-flow graph. Some systems support both pull and push: data flow inspired array fusion~\cite{lippmeier2013data} allows both splits and joins but only for a limited, predefined set of operators. More recent work on polarized data flow fusion~\cite{lippmeier2016polarized} \emph{is} able to fuse our example, but requires the program to be rewritten to use explicitly polarized stream types.

@@

Synchronous dataflow languages such as Lucy-n~\cite{mandel2010lucy} reject value-dependent operators such as @merge@, while general dataflow languages fall back on less performant dynamic scheduling for these cases \cite{bouakaz2013real}. The polyhedral array fusion model~\cite{feautrier2011polyhedron} is used for loop transformations in imperative programs, but operates at a much lower level. The polyhedral model is based around affine loops, which makes it difficult to support for filter-like operators such as @group@ and @merge@.

In our new system we still view the program as a concurrent process network. Each operator is a separate process, and the stream data flows through communication channels between the processes. Each operator is expressed as a restricted, sequential imperative program with commands that include both @pull@ for reading from an input stream and @push@ for writing to an output stream. The fusion transform takes the concurrent process network and \emph{sequentializes} it into a single process by choosing a particular evaluation order that requires no unbounded intermediate buffers. When the fusion transformation succeeds we know it has worked. There is no need to inspect intermediate representations of the compiler to debug poor performance, which is a common problem in systems based on general purpose program transformations \cite{lippmeier2012:guiding}.

### 2017-02-24 2017mergingmerges/section/XX-Processes.tex
Fixing start of S2, also pagination

%!TEX root = ../Main.tex
\section{Processes}
\label{s:Processes}

Each combinator defines a stream process which is expressed in terms of a simple imperative program with a local heap.
The process pulls data from an arbitrary number of input streams and pushes data to at least one output stream.
Each input array is converted to its own process with a single output stream, which repeatedly pushes the values from the array.
% \ben{Mention how we're going to convert arrays to streams, but we don't need to give the details yet}.

% make it explicit that the processes are only an intermediate repr
The process language is used for the implementation of combinators and as an intermediate representation for fusion, while the original program using these combinators can be written without any knowledge of processes.

% -----------------------------------------------------------------------------
\subsection{Grouping}
\begin{figure}

\begin{center}
\begin{alltt}
           group 
             = \(\lambda\) (s1: Stream Nat) (s2: Stream Nat). 
               \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
\end{alltt}
\begin{code}
               process
               { ins:    { s1 }
               , outs:   { s2 }
               , heap:   { f = T, l = 0, v = 0 }
               , label:  A0
               , instrs: { A0 = pull s1 v            A1 {}
                         , A1 = case (f || (l /= v)) A2 {}  A3 {}
                         , A2 = push s2 v            A3 { l = v, f = F }
                         , A3 = drop s1              A0 {} } }
\end{code}
\end{center}
\vspace{1em}
\includegraphics[scale=1.1]{figures/state-group.pdf}
\caption{The group combinator}
\label{fig:Process:Group}
\end{figure}

The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Fig.~\ref{fig:Process:Group}. We include the concrete code representation as well as a diagram of the state machine defined by the associated stream process.

@@

The initial instruction @(pull s1 v A1 {})@ pulls the next element from the stream @s1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@.
The empty set @{}@ after the target label @A1@ can be used to update values in the heap, but as we do not need to perform heap updates here we leave it empty.

@@

When the above predicate is true, the instruction @(push s2 v A3 { l = v, f = F })@ pushes the value @v@ to the stream @s2@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).

Finally, the instruction @(drop s1 A0 {})@ signals that the current element that was pulled from stream @s1@ is no longer required, and goes back to the first the instruction at @A0@.
This @drop@ instruction is used to coordinate processes when performing fusion and concurrent evaluation; the next element of a stream may only be pulled after all consuming processes have pulled and dropped the current element.
% This @drop@ instruction is a synchronisation primitive that is used in the fusion algorithm and concurrent evaluation, but has no direct operational meaning.
% We discuss @drop@ further in \TODO{ref}.

Overall, the @f@ variable tracks whether we are dealing with the first value from the stream, @l@ holds the last value pulled from the stream (or 0 if none have been read yet), and @v@ holds the current value pulled from the stream.
The process emits the first value pulled from the stream and every value that is different from the last one that was pulled.
For example, when executed on the input stream $[1, 2, 2, 3]$, this process will produce the output $[1, 2, 3]$.

Note that we refer to @group@ itself as a \emph{combinator} as it is parameterised by the names of the streams it deals with (@s1@ and @s2@). Once the combinator has been instantiated to work on particular streams the result is a concrete \emph{process}, which is the code representation of a state machine.

% Recall that @group@ removes consecutive duplicates from its input stream.
% It has one input, @file1@, and one output, @uniques@.

% There are three variables in the heap: @first@ is initialised to @True@ as the first pulled value has no last value to compare against; @value@ stores the most recently pulled value, and @last@ stores the last pulled value.

% The @last@ and @value@ variables are initialised to $0$, but could be initialised to any value: these initial values will not be used.
% The initial label is @L0@, which pulls from the input @file1@. This blocks waiting for an input value, and when one is received, it is stored in @value@ and the process moves to @L1@.
% Instruction @L1@ performs a case analysis on a boolean: if it is the first read value, or the last value is not equal to the most recent value, it jumps to @L2@; otherwise it jumps to @L3@.
% Instruction @L2@ pushes the most recent value to the output and jumps to @L3@, updating the last value to the most recent value, and setting first to @False@.
% Finally, the instruction for @L3@ `drops' the recently pulled value from @file1@ and jumps back to @L0@.
% This dropping is required to coordinate between multiple processes that read from the same input: now that the read value from @file1@ has been dropped, another process is free to pull the next value from @file1@ if it so wishes.

% -----------------------------------------------------------------------------
\subsection{Merging}
\begin{figure}
\begin{alltt}
               merge
                 = \(\lambda\) (s1: Stream Nat) (s2: Stream Nat) (s3: Stream Nat). 
                   \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
\end{alltt}
\begin{code}
                   process
                   { ins:    { s1, s2 }
                   , outs:   { s3 }
                   , heap:   { x1 = 0, x2 = 0 }
                   , label:  B0
                   , instrs: { B0 = pull s1 x1     B1 {}
                             , B1 = pull s2 x2     C0 {}
                             , C0 = case (x1 < x2) D0 {}  E0 {}
                             , D0 = push s3 x1     D1 {}
                             , D1 = drop s1        D2 {}
                             , D2 = pull s1 x1     C0 {}
                             , E0 = push s3 x2     E1 {}
                             , E1 = drop s2        E2 {}
                             , E2 = pull s2 x2     C0 {} } }
\end{code}

@@

The definition of the @merge@ combinator, which merges two input streams, is given in Fig.~\ref{fig:Process:Merge}.
The combinator binds the two input streams to @s1@ and @s2@, while the output stream is @s3@.
The two heap variables @x1@ and @x2@ are used to store the currently read values from each input.
The process starts by pulling from both input streams, then compares the two values, and pushes the smaller of the values to the output stream.
It then drops the input stream corresponding to the smaller value, then pulls from the same stream, and performs the same comparison again.
% The forms of the instructions used to define this combinator are the same as the @group@ combinator in the previous section.
% We only need the four basic @pull@, @case@, @push@ and @drop@ instructions.

@@

% The instructions at @L0@ and @L1@ initialise the process by reading the first values from each stream, then move to @C0@.
% Instruction @C0@ checks which value is smaller: if the value @a@ read from @file1@ is smaller, it moves to @A0@; otherwise it moves to @B0@.
% Instructions @A0@, @A1@ and @A2@ push @file1@'s value to the output, drop it, and pull another one before moving back to @C0@ to compare again.
% Instructions @B0@, @B1@ and @B2@ are equivalent, except performing on @file2@ instead.

% -----------------------------------------------------------------------------
\clearpage{}
\subsection{Fusion}
Our fusion algorithm takes two process state machines and produces a new one that computes the outputs of both.
For example, suppose we need to produce a single machine that computes the outputs of the first two combinators of our @uniquesUnion@ example back in \S\ref{s:Introduction}.
We want a result machine that computes both a @group@ machine and a @merge@ machine executed in parallel, where the first input stream of the @merge@ is the same as the input stream of the @group@.
Compared to the @uniquesUnion@ example, we will associate a stream named @s1@ with the first array (@arr1@) and a stream named @s2@ with the second array (@arr2@).

% -----------------------------------------------------------------------------
\subsubsection{Fusing Pulls}
\begin{figure}
\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
\caption{Fusing pull instructions}
\label{fig:Fusion:Pulls}
\end{figure}

@@

We also associate each of the result states with information describing whether or not each machine has already pulled a value from each input stream, which is used to determine whether another value can be pulled.
For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{s1 = none\}), (B0, \{s1 = none, s2 = none\})).
This means that the result state @F0@ represents a combination of the two input states @A0@ and @B0@.
As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both @s1@ and @s2@ are set to `none'.

From the joint state @F0@, both of the input machines need to pull from stream @s1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@.
In the result machine this is managed by storing the pulled value in a fresh buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@.
For this we use the @jump@ instruction which transitions between states without affecting any of the input or output streams.
After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.

@@

\eject{}
% -----------------------------------------------------------------------------
\subsubsection{Fusing Cases}
\begin{figure}
\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
\caption{Fusing case instructions}
\label{fig:Fusion:Case}
\end{figure}

@@

\clearpage{}
\subsection{Process definitions}

>>>>

%!TEX root = ../Main.tex
\section{Processes, Combinators and Operators}
\label{s:Processes}

A \emph{process} in our system is a simple imperative program with a local heap. A process pulls data from an arbitrary number of input streams and pushes result values to at least one output stream. The process language is an intermediate representation we use when fusing the overall dataflow network. A \emph{combinator} is a template for a particular \emph{process} which parameterises it over the particular input and output streams, as well as values of configuration parameters such as the worker function used in a @map@ process. Each process implements a logical \emph{operator}, so we use ``operator'' when describing the values being computed, but ``process'' when referring to the implementation.

% -----------------------------------------------------------------------------
\subsection{Grouping}

The definition of the @group@ combinator which removes consecutive elements from its input stream is given in Fig.~\ref{fig:Process:Group}. We include the concrete code representation and a diagram of the process viewed as a state machine.

@@

The initial instruction @(pull s1 v A1 {})@ pulls the next element from the stream @s1@, writes it into the heap variable @v@ (value), then proceeds to the instruction at label @A1@. The empty set @{}@ after the target label @A1@ can be used to update values in the heap, but as we do not need to perform heap updates yet we leave it empty.

@@

When the predicate is true, the instruction @(push s2 v A3 { l = v, f = F })@ pushes the value @v@ to the output stream @s2@ and proceeds to the instruction at label @A3@, once the heap has been updated to set variable @l@ to @v@ and @f@ to @F@ (False).

Finally, the instruction @(drop s1 A0 {})@ signals that the current element that was pulled from stream @s1@ is no longer required, and goes back to the first the instruction at @A0@. This @drop@ instruction is used to coordinate concurrent processes when performing fusion. The next element of a stream may only be pulled after all consuming processes have pulled and then and dropped the current element.

Overall, the @f@ variable tracks whether we are dealing with the first value from the stream, @l@ holds the last value pulled from the stream (or 0 if none have been read yet), and @v@ holds the current value pulled from the stream. The process emits the first value pulled from the stream and every value that is different from the last one that was pulled. For example, when executed on the input stream $[1, 2, 2, 3]$, the process will produce the output $[1, 2, 3]$.

\begin{figure}

\begin{center}
\begin{alltt}
           group 
             = \(\lambda\) (sG1: Stream Nat) (sG2: Stream Nat). 
               \(\nu\) (f: Bool) (l: Nat) (v: Nat) (A0..A3: Label).
\end{alltt}
\begin{code}
               process
               { ins:    { sG1 }
               , outs:   { sG2 }
               , heap:   { f = T, l = 0, v = 0 }
               , label:  A0
               , instrs: { A0 = pull sG1 v           A1 {}
                         , A1 = case (f || (l /= v)) A2 {}  A3 {}
                         , A2 = push sG2 v           A3 { l = v, f = F }
                         , A3 = drop sG1             A0 {} } }
\end{code}
\end{center}
\vspace{1em}
\includegraphics[scale=1.1]{figures/state-group.pdf}
\caption{The group combinator}
\label{fig:Process:Group}
\end{figure}

% -----------------------------------------------------------------------------
\subsection{Merging}
\begin{figure}
\begin{alltt}
               merge
                 = \(\lambda\) (sM1: Stream Nat) (sM2: Stream Nat) (sM3: Stream Nat). 
                   \(\nu\) (x1: Nat) (x2: Nat) (B0..E2: Label).
\end{alltt}
\begin{code}
                   process
                   { ins:    { s1, s2 }
                   , outs:   { s3 }
                   , heap:   { x1 = 0, x2 = 0 }
                   , label:  B0
                   , instrs: { B0 = pull s1 x1     B1 {}
                             , B1 = pull s2 x2     C0 {}
                             , C0 = case (x1 < x2) D0 {}  E0 {}
                             , D0 = push s3 x1     D1 {}
                             , D1 = drop s1        D2 {}
                             , D2 = pull s1 x1     C0 {}
                             , E0 = push s3 x2     E1 {}
                             , E1 = drop s2        E2 {}
                             , E2 = pull s2 x2     C0 {} } }
\end{code}

@@

The definition of the @merge@ combinator, which merges two input streams, is given in Fig.~\ref{fig:Process:Merge}. The combinator binds the two input streams to @s1@ and @s2@, while the output stream is @s3@. The two heap variables @x1@ and @x2@ are used to store the last values read from each input stream. The process starts by pulling from each input streams. It then compares the two pulled values, and pushes the smaller of the values to the output stream. The process then drops the stream which yielded the the smaller value, then pulls from the same stream so that it can perform the comparison again.

@@

% -----------------------------------------------------------------------------
\subsection{Fusion}
Our fusion algorithm takes two process state machines and produces a new one that computes the outputs of both. For example, suppose we need a single machine that computes the outputs of the first two combinators of our @uniquesUnion@ example back in \S\ref{s:Introduction}. The result will be a machine that computes the result of both the @group@ and @merge@ if they were executed concurrently, where the first input stream of the @merge@ is the same as the input stream of the @group@. For ease of comparison we will assume the parameters in each combinator are instantiated with arguments of the same names.

% -----------------------------------------------------------------------------
\subsubsection{Fusing Pulls}

@@

We also associate each of the result states with information describing whether or not each machine has already pulled a value from each input stream, which is used to determine whether another value can be pulled. For the @F0@ case shown in Fig.~\ref{fig:Fusion:Pulls} we have ((A0, \{s1 = none\}), (B0, \{s1 = none, s2 = none\})). The result state @F0@ represents a combination of the two input states @A0@ and @B0@. As both @A0@ and @B0@ are the initial states of their respective machines, those machines have not yet pulled any values from their two input streams, so both @s1@ and @s2@ are set to `none'.

From the joint state @F0@, both of the input machines need to pull from stream @s1@, the @group@ machine storing the value in a variable @v@ and the @merge@ machine storing it in @x1@.
In the result machine this is managed by storing the pulled value in a fresh buffer variable @b1@, and then using later instructions to copy the value into the original variables @v@ and @x1@.
For this we use the @jump@ instruction which transitions between states without affecting any of the input or output streams.

@@

\begin{figure}
\includegraphics[scale=1.1]{figures/fuse-pull-pull.pdf}
\caption{Fusing pull instructions}
\label{fig:Fusion:Pulls}
\end{figure}

% -----------------------------------------------------------------------------
\subsubsection{Fusing Cases}
\begin{figure}
\includegraphics[scale=1.1]{figures/fuse-case-pull.pdf}
\caption{Fusing case instructions}
\label{fig:Fusion:Case}
\end{figure}

@@

\subsection{Optimisation}
After the algorithm has completed fusing both processes it may be possible to eliminate redundant @jump@, depending on how the affected variables are used by other instructions.

\TODO{Discuss combining jump instructions here. Use this to motivate the need for drop instructions.}

% ----------------------------------------------------------------------------------
\clearpage{}
\subsection{Process definitions}
